### 2025-05-06 lecturenotes_aima203/ch2_aima203.tex
python code bisection

\section{Iteration Method}
Steps:
\begin{enumerate}
\item Re-write the given equation \(f(x)=0\) in the form \(x= \phi (x)\).
  This equation is of \textbf{iterative-type}. Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\) if the iteration is of convergent one.
\item Choose an initial root of \(f\), \(x_0\).
\item \(x_1=\phi (x_0)\), \(x_2= \phi(x_1)\) and so on.
\end{enumerate}
The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number. But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=11cm]{secant_example.png}
\end{figure}

\section{System of Non-linear equations}
For now we consider only a system of two equations. Let a system of two equations be
\begin{equation}
  \label{system}
  f(x,y)=0, \hspace{1cm} g(x,y)=0
\end{equation}
\subsection{Method of Iteration}
First we assume that the sytem of equations~\ref{system} may be written in the form
\begin{equation}
  \label{iterate}
  x=F(x,y), \hspace{1cm} y=G(x,y)
\end{equation}
where the function $F$ and $G$ satisty the following conditions in a \textbf{closed} neighborhood of $R$ of the root $(\alpha, \beta)$:
\begin{enumerate}
\item[i)] $F$ and $G$ and their firt partial derivatives are continuous in $R$, and
  \item[ii)] $\displaystyle \left | \frac{\partial F}{\partial x} \right | + \left | \frac{\partial F}{\partial y} \right | < 1$ and $\displaystyle \left | \frac{\partial G}{\partial x} \right | + \left | \frac{\partial G}{\partial y} \right | < 1$, for all $(x,y)$ in $R$.
\end{enumerate}
If $(x_0,y_0)$ is an initial approximation to the root $(\alpha, \beta)$, then Equations~\ref{iterate} give the sequence
\begin{equation}
\begin{gathered}
  x_1 = F(x_0,y_0), \hspace{2cm} y_1 = G(x_0,y_0) \\
  x_2 = F(x_1,y_1), \hspace{2cm} y_2 = G(x_1,y_1) \\
  ... \\
  x_{n+1} = F(x_n,y_n), \hspace{2cm} y_{n+1} = G(x_n,y_n)
\end{gathered}
\end{equation}
For faster convergence, recently computed values of $x_i$ may be used in the evaluation of $y_i$ in Equations. Above conditions are sufficient for convergence and in the limit we obtain,
\begin{equation}
  \alpha = F(\alpha, \beta) \hspace{5mm} and \hspace{1cm} \beta=G(\alpha, \beta)
\end{equation}
Hence $(\alpha, \beta)$ is the root of the system~\ref{system}.

\subsection{Newton-Raphson Method}
Let $(x_0,y_0$ be an initial approximation to the root of the system of equations in two variables~\ref{system}. If $(x_0+h, y_0+k$ is the root of the system, then we must have
\begin{equation*}
  f(x_0+h,y_0+k) = 0 \hspace{1cm} g(x_0+h,y_0+k) = 0
\end{equation*}
Assuming that $f$ and $g$ are sufficiently differentiable, we expand both of these functions by Taylor's series to obtain
\begin{equation*}
  \begin{gathered}[t]
    f_0 +h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = 0 \\
    g_0 +h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = 0
  \end{gathered}
\end{equation*}
where, \hspace{5mm} $\displaystyle \frac{\partial f}{\partial x_0} = \left [\frac{\partial f}{\partial x} \right ] _{x=x_0}$, $f_0 = f(x_0,y_0)$, etc \\[1mm]
Neglating the second and higher-order derivatives terms, we get,
\begin{equation}
  \label{taylor}
  \begin{gathered}[b]
    h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = -f_0 \\
     h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = g_0
  \end{gathered}
\end{equation}
The system of equations~\ref{taylor} possesses a unique solution if
$$ D =
\left \vert \begin{matrix}
  \frac{\partial f}{\partial x_0}  & \frac{\partial f}{\partial y_0} \\[2mm]
  \frac{\partial g}{\partial x_0}  & \frac{\partial g}{\partial y_0}
\end{matrix} \right \vert  \neq 0
$$
By Cramer's rule
\begin{equation}
  h= \frac{1}{D}\, \left \vert \begin{matrix}
  -f_0  & \frac{\partial f}{\partial y_0} \\[2mm]
  -g_0  & \frac{\partial g}{\partial y_0}
\end{matrix} \right \vert \hspace{5mm} and \hspace{5mm} k= \frac{1}{D}\, \left \vert \begin{matrix}
  \frac{\partial f}{\partial y_0} & -f_0  \\[2mm]
 \frac{\partial g}{\partial y_0} &  -g_0
\end{matrix} \right \vert
\end{equation}
The new approximations are, therefore
\begin{equation}
  x_1 = x_0 +h \hspace{5mm} and \hspace{5mm} y_1 = y_0 + k
\end{equation}

@@

@@

\section{Lab Work}
\begin{enumerate}
\item Evaluation of Polynomials
\end{enumerate}

>>>>

\section{Iteration Method}
Steps:
\begin{enumerate}
\item Re-write the given equation \(f(x)=0\) in the form \(x= \phi (x)\).
  This equation is of \textbf{iterative-type}. Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\), if the iteration is of convergent one.
\item Choose an initial root of \(f\), \(x_0\).
\item \(x_1=\phi (x_0)\), \(x_2= \phi(x_1)\) and so on.
\end{enumerate}
The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number. But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation.

@@

\section{System of Non-linear equations}
For now we consider only a system of two equations. Let a system of two equations be
\begin{equation}
  \label{system}
  f(x,y)=0, \hspace{1cm} g(x,y)=0
\end{equation}
\subsection{Method of Iteration}
First we assume that the sytem of equations~\ref{system} may be written in the form
\begin{equation}
  \label{iterate}
  x=F(x,y), \hspace{1cm} y=G(x,y)
\end{equation}
where the function $F$ and $G$ satisty the following conditions in a \textbf{closed} neighborhood of $R$ of the root $(\alpha, \beta)$:
\begin{enumerate}
\item[i)] $F$ and $G$ and their firt partial derivatives are continuous in $R$, and
  \item[ii)] $\displaystyle \left | \frac{\partial F}{\partial x} \right | + \left | \frac{\partial F}{\partial y} \right | < 1$ and $\displaystyle \left | \frac{\partial G}{\partial x} \right | + \left | \frac{\partial G}{\partial y} \right | < 1$, for all $(x,y)$ in $R$.
\end{enumerate}
If $(x_0,y_0)$ is an initial approximation to the root $(\alpha, \beta)$, then Equations~\ref{iterate} give the sequence
\begin{equation}
\begin{gathered}
  x_1 = F(x_0,y_0), \hspace{2cm} y_1 = G(x_0,y_0) \\
  x_2 = F(x_1,y_1), \hspace{2cm} y_2 = G(x_1,y_1) \\
  ... \\
  x_{n+1} = F(x_n,y_n), \hspace{2cm} y_{n+1} = G(x_n,y_n)
\end{gathered}
\end{equation}
For faster convergence, recently computed values of $x_i$ may be used in the evaluation of $y_i$ in Equations. Above conditions are sufficient for convergence and in the limit we obtain,
\begin{equation}
  \alpha = F(\alpha, \beta) \hspace{5mm} and \hspace{1cm} \beta=G(\alpha, \beta)
\end{equation}
Hence $(\alpha, \beta)$ is the root of the system~\ref{system}.
\begin{figure}[h]
  \centering
\includegraphics[width=13cm, height=11cm]{secant_example.png}
\end{figure}

\subsection{Newton-Raphson Method}
Let $(x_0,y_0$ be an initial approximation to the root of the system of equations in two variables~\ref{system}. If $(x_0+h, y_0+k$ is the root of the system, then we must have
\begin{equation*}
  f(x_0+h,y_0+k) = 0 \hspace{1cm} g(x_0+h,y_0+k) = 0
\end{equation*}
Assuming that $f$ and $g$ are sufficiently differentiable, we expand both of these functions by Taylor's series to obtain
\begin{equation*}
  \begin{gathered}[t]
    f_0 +h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = 0 \\
    g_0 +h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = 0
  \end{gathered}
\end{equation*}
where, \hspace{5mm} $\displaystyle \frac{\partial f}{\partial x_0} = \left [\frac{\partial f}{\partial x} \right ] _{x=x_0}$, $f_0 = f(x_0,y_0)$, etc \\[1mm]
Neglating the second and higher-order derivatives terms, we get,
\begin{equation}
  \label{taylor}
  \begin{gathered}[b]
    h\frac{\partial f}{\partial x_0} + k\frac{\partial f}{\partial y_0}... = -f_0 \\
     h\frac{\partial g}{\partial x_0} + k\frac{\partial g}{\partial y_0}... = g_0
  \end{gathered}
\end{equation}
The system of equations~\ref{taylor} possesses a unique solution if
$$ D =
\left \vert \begin{matrix}
  \frac{\partial f}{\partial x_0}  & \frac{\partial f}{\partial y_0} \\[2mm]
  \frac{\partial g}{\partial x_0}  & \frac{\partial g}{\partial y_0}
\end{matrix} \right \vert  \neq 0
$$
By Cramer's rule
\begin{equation}
  h= \frac{1}{D}\, \left \vert \begin{matrix}
  -f_0  & \frac{\partial f}{\partial y_0} \\[2mm]
  -g_0  & \frac{\partial g}{\partial y_0}
\end{matrix} \right \vert \hspace{5mm} and \hspace{5mm} k= \frac{1}{D}\, \left \vert \begin{matrix}
  \frac{\partial f}{\partial y_0} & -f_0  \\[2mm]
 \frac{\partial g}{\partial y_0} &  -g_0
\end{matrix} \right \vert
\end{equation}
The new approximations are, therefore
\begin{equation}
  x_1 = x_0 +h \hspace{5mm} and \hspace{5mm} y_1 = y_0 + k
\end{equation}
\begin{figure}[h]
  \centering
\includegraphics[width=15cm, height=7cm]{newton_example.png}
\end{figure}

@@

\section{Rate of Convergence}
If a sequence $x_1, x_2, x_3, ...$ converges to a value $\alpha$ then, the rate of convergence of the sequence $(x_n)$ to $\alpha$ is $p$, if $p$ is the \textbf{largest} possible number such that:
\begin{equation}
  \label{rateofconverge}
 \lim_{n \to \infty} \;  \frac{|x_{n+1} - \alpha|}{|x_n-\alpha|^p} = constant
\end{equation}
In another words, if $\epsilon _n = x_n-\alpha$ is the error at $nth$ step then, the sequence of the error $(\epsilon_n)$ is a decreasing sequence and the positive integer $p$ defines the rate at which the sequence of error is decreasing and converging to the correct value $\alpha$. So we have,
 e\begin{equation*}
 \lim_{n \to \infty} \;  \frac{|\epsilon _{n+1}|}{|\epsilon _n|^p} = constant
\end{equation*}
When $p=1$, we say the sequence or the method converges \textbf{linearly}, when $p=2$ we say the method converges \textbf{quadratically}, when $1 < p < 2$, we say the method converges \textbf{superlinearly}.

\section{Exercise}
\begin{enumerate}
\item Find the rate of convergence of Bisection Method.
\item The rate of convergence of Newton Rapshon's method is quadratic. Does it converge faster than Bisection Method?
\item Write the rate of convergence of secant method.
\end{enumerate}

\section{Lab Work}
\begin{enumerate}
\item Try a suitable data structure for storing the sequence of outputs in a iteration.
\item Find a way of representing a function.
\item Find an efficient way of evaluating a function or a polynomial.
\item Draw a flow chart for the working of Bisection Method.
\item Writing a function in Python.
\end{enumerate}

\begin{lstlisting}
  def bisection_method(f, a, b, tol=1e-6, max_iter=100):
    """
    Find a root of a function f(x) in the interval [a, b] using the bisection method.

    Parameters:
    f (function): The function for which to find a root.
    a (float): Left endpoint of the interval.
    b (float): Right endpoint of the interval.
    tol (float): Tolerance (stopping criterion).
    max_iter (int): Maximum number of iterations.

    Returns:
    float: Approximation of the root.
    int: Number of iterations performed.
    """
    # Check if the interval contains a root
    if f(a) * f(b) >= 0:
    raise ValueError("Function must have opposite signs at the
    endpoints of the interval.")

    iter_count = 0
    c = a

    while (b - a) / 2 > tol and iter_count < max_iter:
        c = (a + b) / 2  # Midpoint

        if f(c) == 0:
            break  # Exact solution found
        elif f(a) * f(c) < 0:
            b = c  # Root is in left half
        else:
            a = c  # Root is in right half

        iter_count += 1

        return c, iter_count

# Example usage:
if __name__ == "__main__":
    # Define the function whose root we want to find
    def f(x):
        return x**3 - 2*x - 5

    # Initial interval
    a, b = 2, 3

    # Apply bisection method
    root, iterations = bisection_method(f, a, b)

    print(f"Approximate root: {root:.6f}")
    print(f"Found in {iterations} iterations")
    print(f"f(root) = {f(root):.6f}")
\end{lstlisting}

@@
